---
title: "Initial Analysis"
author: "Lindsey Ragsdale"
date: today
editor: visual
editor_options: 
  chunk_output_type: console
format: docx
execute: 
  echo: FALSE
  message: FALSE
---

# **Where is this data from?**

*For this project I used a .csv data files from IPEDS survey year 2022, Graduation rate data for Pell Grant and Direct Subsidized Loan recipients, 150% of normal time to complete - cohort year 2016 (4-year) and cohort year 2019 (less-than-4-year) institutions.* This is IPEDS complete data that can be found here: <https://nces.ed.gov/ipeds/datacenter/data/GR2022_PELL_SSL.zip>. *To run this code, save this file in a sub-folder called “data” that sits in the same folder as this .qmd file.*

### Load Library

```{r}
library(tidyverse)
library(haven)
library(dtplyr)
library(ggplot2)
library(knitr)
library(patchwork)

```

### Read Data

```{r}
df<-read_csv(file.path("data", "gr2022_pell_ssl.csv"))
```

### #Select columns to work with. Best practice to keep UNITID, then compare

```{r}
df_pvt <- df |>
  select(UNITID, PSGRTYPE, PGADJCT, NRADJCT, PGCMTOT, TTCMTOT) 
```

filter out 2 and 3 from PSGRTYPE

```{r}
gdf_pvt <- df_pvt |>
  filter(PSGRTYPE %in% c(1, 4))

```

#create percentages: #1. Pell graduates/ total *100 = PGCMTOT/(PGADJCT+NRADJCT)* 100 #2. Total graduates/ total *100 = TTCMTOT/(PGADJCT+NRADJCT)* 100

```{r}
df_pvt <- df_pvt |>
  mutate(Pell_percent = (PGCMTOT / (PGADJCT + NRADJCT)) * 100)

print(df_pvt)

df_pvt <- df_pvt |>
  mutate(NonPell_percent = (TTCMTOT / (PGADJCT + NRADJCT)) * 100)

print (df_pvt)

```

Histogram showing Pell v. Non-Pell graduates

```{r}
df_percent <- df_pvt |> 
  pivot_longer(cols = c(Pell_percent, NonPell_percent),
               names_to = "percentage_type",
               values_to = "percentage_value")
ggplot(df_percent) +
  geom_histogram(aes(x = percentage_value,
                     fill = percentage_type),
                 alpha = 0.5,
                 color = "black")

    labs(title = "Percentage of Pell Grant and Non-Pell Grant Graduates",
         subtitle = NULL,
         caption = NULL,
         x = "Percentage",
         y = "Fequency") +
    theme_minimal()
  
    
```

Box plot

```{r}
df_box <- df_pvt |> 
      pivot_longer(cols = c(Pell_percent, NonPell_percent),
                   names_to = "percentage_type",
                   values_to = "percentage_value")  
    
ggplot(df_box)+
  geom_boxplot(mapping= aes(x= percentage_type,
                            y= percentage_value,
                            fill=percentage_type ), 
               alpha = 0.66) 
```

Scatter Plot

```{r}
ggplot(data = df_pvt,
       mapping = aes(x = Pell_percent,
                     y = NonPell_percent)) +
  geom_point() +
  labs(title = "Scatter Plot of Pell Graduates vs Total Graduates",
       x = "Pell Graduates Percentage",
       y = "Non-Pell Graduates Percentage")
```
